Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression

### 2024-11-30 paper/paper.qmd
comments updated, EDA re-structured
comments updated, EDA re-structured

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
set.seed(912)
# List of necessary packages
packages <- c("tidyverse", "dplyr", "arrow", "ggplot2", "RColorBrewer", "knitr", "bayesplot", "kableExtra")
# Install missing packages
missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
install.packages(missing_packages)
# Load packages
lapply(packages, library, character.only = TRUE)

@@

## Measurement

To predict the relationship between bike-sharing usage and time within the University of Toronto's St. George campus, this study requires a time series to describe the changes in usage at different stations over time. Therefore, the "Bike Share Toronto Ridership Data" [@bikesharedata], which spans over eight years to date and documents every instance of bike-sharing usage, including the time and location of each ride, is ideal for this study.

@@

However, the raw data cannot be used directly in this project; sophisticated data cleaning is required, with specific steps and reasons as follows: First, data from 2015-2016 was excluded because the collection and recording methods for those years differ significantly from later years and do not include the time and station variables required for this study. Next, data for stations located within the University of Toronto's St. George campus was extracted from all remaining samples, and the usage of each station was calculated for every four-hour interval. Finally, the variables of interest were extracted, and the data format was standardized for subsequent analysis. The cleaned data is presented in the format of @tbl-cleaned_data.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleaned_data
#| tbl-cap: "Samples of the dataset used for analysis"
start |>
  tail() |>
  kable(booktabs = TRUE, align = "c")
```

## Variables

This study focuses on the following variables:

- `count`: The dependent variable of the study, a non-negative integer. It describes the total usage of bike-sharing at a particular station during a specific 4-hour interval.
- `time`: An independent variable representing the time interval. For example, "2024-09-29 12:00:00" represents the time interval from 12 pm to 4 pm on September 29, 2024, in a 24-hour format. The earliest `time` is "2017-01-01 00:00:00," and the latest is "2024-09-30 24:00:00."
- `station_name`: An independent variable representing the unique name of one of the 27 stations within the University of Toronto's St. George campus.

@@

ggplot(data = daily_counts, aes(x = time, y = total_count)) +
  geom_point() +
  labs(x = "Time", y = "Count") +
  theme_minimal() +
  geom_smooth(method = "lm") +
  annotate("text", x = max(daily_counts$time), y = max(daily_counts$total_count), 
           label = paste("Slope =", round(slope, 2)), hjust = 1.3)

@@

hows the total usage and percentage of shared bicycles during different time intervals in a day from January 1, 2017, to September 30, 2024. It can be observed that the peak usage occurs between 12:00 to 16:00 (28.5%) and 16:00 to 20:00 (31.4%). The usage between 08:00 to 12:00 accounts for 19.9%, while the usage from 20:00 to 00:00 accounts for 13.4%. The time intervals with the lowest usage are 00:00 to 04:00 (3.6%) and 04:00 to 08:00 (3.2%).

```{r, fig.pos="H"}
#| label: fig-timely_usage
#| fig-cap: Total usage during different time periods from January 2017 to September 2024
#| echo: false
#| message: false
#| warning: false
options(scipen = 999)

@@

ggplot(data = total_counts, aes(x = time_interval, y = count, fill = time_interval)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 3) + 
  labs(x = "Time interval", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

@@

# Model {#sec-model}

## Model descriptions
The model used in this study is as follows:

\begin{equation*}
\begin{aligned}
    \text{count} &\sim \text{Poisson}(\lambda) \\
    \log(\lambda) &= \beta_0 + \beta_1 \times \text{hour} + \beta_2 \times \text{day} + \beta_3 \times \text{month} + \beta_4 \times \text{year} \\
    \beta_0 &\sim \mathcal{N}(0, 5) \\
    \beta_j &\sim \mathcal{N}(0, 2) \quad \text{for } j = 1, 2, 3, 4
\end{aligned}
\end{equation*}

@@

This study considered using the ARIMA (autoregressive integrated moving average) model as an alternative but ultimately decided against it due to its limitations in capturing the specific characteristics of bike-sharing usage data. Bike-sharing usage exhibits significant seasonal patterns, time-of-day effects, and count-based characteristics, while ARIMA models are primarily suited for stationary time series with linear autocorrelations. ARIMA struggles to effectively address non-stationary trends and overdispersion, such as the marked decrease in usage during winter. Although extensions like SARIMA (Seasonal ARIMA) can account for seasonality, they still cannot directly handle the count-based nature of the data. In contrast, Bayesian Poisson regression is specifically designed for count data, allowing for the flexible integration of variables such as year, month, and specific four-hour intervals. Additionally, the Bayesian framework provides posterior distributions, offering deeper insights into parameter uncertainty and variability, which are critical for policy decisions. Therefore, Bayesian Poisson regression is more suitable for this study than the ARMA model.

@@

```{r, fig.pos="H"}
#| label: fig-PPplot
#| fig-cap: Posterior Predictive Checks
#| echo: false
#| message: false
#| warning: false
# 5. Posterior Predictive Checks
pp_check(poisson_glm, plotfun = "dens_overlay", nreps = 100) +
  labs(
    x = "Count",  
    y = "Estimated density"         
  ) 
```
The credible intervals shown in @tbl-crediinte highlight several strengths of the model. Key variables such as hour, month, and year have narrow credible intervals that do not include zero, indicating their significant and consistent influence on the response variable. This demonstrates the model’s ability to effectively capture critical trends and relationships in the data, particularly those related to time and seasonal effects. Additionally, the relatively small width of these intervals reflects the stability and reliability of parameter estimates, suggesting that the model is well-calibrated and robust in its predictions. The intercept, while having a broader credible interval, provides a reasonable baseline estimate, reflecting the overall level of the response variable.

@@

# Create the plot
ggplot(data_long, aes(x = station_name, y = value, fill = Note)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(aes(label = value), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, 
            size = 3) + 
  labs(
    x = "Station Name",
    y = "Count",
  ) +
  scale_fill_manual(
    values = c("count" = "pink", "prediction" = "skyblue"),
    labels = c("count" = "Actual count in 2024", "prediction" = "Predicted count in 2025") 
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

>>>>

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
set.seed(912)
# List of necessary packages
packages <- c("tidyverse", "dplyr", "arrow", "ggplot2", "RColorBrewer", "knitr", "bayesplot", "kableExtra", "scales")
# Install missing packages
missing_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
install.packages(missing_packages)
# Load packages
lapply(packages, library, character.only = TRUE)

@@

**scales**[@scales]: Used for formatting large numbers.

## Measurement
To predict the relationship between bike-sharing usage and time within the University of Toronto's St. George campus, this study requires a time series to describe the changes in usage at different stations over time. Therefore, the "Bike Share Toronto Ridership Data" [@bikesharedata], which spans over eight years to date and documents every instance of bike-sharing usage, including the time and location of each ride, is ideal for this study.

@@

However, the raw data cannot be used directly in this project; sophisticated data cleaning is required, with specific steps and reasons as follows: First, data from 2015–2016 was excluded because the collection and recording methods during those years differed significantly from later periods and did not include the time and station variables necessary for this study. Next, data from stations located within the University of Toronto's St. George campus was extracted from the remaining samples to align with the study's objectives. Then, the variables `time` and `count` were constructed to record the total usage at each station within specific four-hour intervals. Finally, the variables of interest were extracted, and the data format was standardized to ensure compatibility for subsequent analysis. The cleaned data is presented in the format shown in @tbl-cleaned_data.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleaned_data
#| tbl-cap: "Samples of the dataset used for analysis"

start |>
  tail() |>
  kable(booktabs = TRUE, align = "c")
```

## Variables
This study focuses on the following variables:

- `count`: The dependent variable of the study, a non-negative integer. It describes the total usage of bike-sharing at a particular station during a specific 4-hour interval. This is a constructed variable derived by calculating the total number of uses at a specific station within a given four-hour interval.
- `time`: An independent variable representing the time interval. For example, "2024-09-29 12:00:00" represents the time interval from 12 pm to 4 pm on September 29, 2024, in a 24-hour format. The earliest `time` is "2017-01-01 00:00:00," and the latest is "2024-09-30 24:00:00". This is a constructed variable designed to indicate which four-hour interval the exact trip start time, recorded to the second in the raw data, falls into.
- `station_name`: An independent variable representing the unique name of one of the 27 stations within the University of Toronto's St. George campus.

It is important to note that there may be a relationship between the two independent variables, `time` and `station_name`, as the operating times of different stations might vary. However, this does not affect the validity of the study because the analysis focuses on the usage patterns within the available operational hours for each station, rather than comparing stations with differing start times directly.

@@

ggplot(data = daily_counts, aes(x = time, y = total_count)) +
  geom_point() +
  labs(x = "Time", y = "Daily usage") +
  theme_minimal() +
  geom_smooth(method = "lm") +
  annotate("text", x = max(daily_counts$time), y = max(daily_counts$total_count), 
           label = paste("Slope =", round(slope, 2)), hjust = 1.3)

@@

@fig-timely_usage shows the total usage and percentage of shared bicycles during different time intervals in a day from January 1, 2017, to September 30, 2024. It can be observed that the peak usage occurs between 12:00 to 16:00 (28.5%) and 16:00 to 20:00 (31.4%). The usage between 08:00 to 12:00 accounts for 19.9%, while the usage from 20:00 to 00:00 accounts for 13.4%. The time intervals with the lowest usage are 00:00 to 04:00 (3.6%) and 04:00 to 08:00 (3.2%).

```{r, fig.pos="H"}
#| label: fig-timely_usage
#| fig-cap: Total usage during different time periods from January 2017 to September 2024
#| echo: false
#| message: false
#| warning: false

options(scipen = 999)

@@

ggplot(data = total_counts, aes(x = time_interval, y = count, fill = time_interval)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            vjust = -0.5, size = 3) +
  labs(x = "Time interval", y = "Total usage") +
  scale_y_continuous(labels = scales::comma) + 
  theme_minimal() +
  theme(legend.position = "none")

@@

# Model {#sec-model}
In the process of model development, this study utilized R [@citeR] and the following R packages:brms[@brms] and rstanarm[@rstanarm] were used for model construction. bayesplot[@bayesplot], tidyverse[@tidyverse], ggplot2[@ggplot2], knitr[@knitr], and kableExtra[@kableExtra] were used for verifying model assumptions and model validation.

## Model descriptions
Equation\ref{eq:poisson} is the model used in this study:

\begin{equation}
\begin{aligned}
    \text{count} &\sim \text{Poisson}(\lambda) \\
    \log(\lambda) &= \beta_0 + \beta_1 \times \text{hour} + \beta_2 \times \text{day} + \beta_3 \times \text{month} + \beta_4 \times \text{year} \\
    \beta_0 &\sim \mathcal{N}(0, 5) \\
    \beta_j &\sim \mathcal{N}(0, 2) \quad \text{for } j = 1, 2, 3, 4
\end{aligned}
\label{eq:poisson}
\end{equation}

@@

This study considered using the ARIMA (autoregressive integrated moving average) model as an alternative but ultimately decided against it due to its limitations in capturing the specific characteristics of bike-sharing usage data. Although the ARIMA model has advantages in time series analysis, allowing for more accurate short-term predictions with parameters that have clearer interpretability. Bike-sharing usage exhibits significant seasonal patterns, time-of-day effects, and count-based characteristics, while ARIMA models are primarily suited for stationary time series with linear autocorrelations. ARIMA struggles to effectively address non-stationary trends and overdispersion, such as the marked decrease in usage during winter. Although extensions like SARIMA (Seasonal ARIMA) can account for seasonality, they still cannot directly handle the count-based nature of the data. In contrast, Bayesian Poisson regression is specifically designed for count data, allowing for the flexible integration of variables such as year, month, and specific four-hour intervals. Additionally, the Bayesian framework provides posterior distributions, offering deeper insights into parameter uncertainty and variability, which are critical for policy decisions. Therefore, Bayesian Poisson regression is more suitable for this study than the ARMA model.

@@

```{r, fig.pos="H"}
#| label: fig-PPplot
#| fig-cap: Posterior Predictive Checks
#| echo: false
#| message: false
#| warning: false
# 5. Posterior Predictive Checks
pp_check(poisson_glm, plotfun = "dens_overlay", nreps = 100) +
  labs(
    x = "Usage",  
    y = "Estimated density"         
  ) +
   yaxis_text()
```
The credible intervals shown in @tbl-crediinte highlight several strengths of the model. Key variables such as hour, month, and year have narrow credible intervals that do not include zero, indicating their significant and consistent influence on the response variable. This demonstrates the model’s ability to effectively capture critical trends and relationships in the data, particularly those related to time and seasonal effects. Additionally, the relatively small width of these intervals reflects the stability and reliability of parameter estimates, suggesting that the model is well-calibrated and robust in its predictions. The intercept, while having a broader credible interval, provides a reasonable baseline estimate, reflecting the overall level of the response variable.

@@

# Create the plot
ggplot(data_long, aes(x = station_name, y = value, fill = Note)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(aes(label = value), 
            position = position_dodge(width = 0.7), 
            vjust = -0.5, 
            size = 3) + 
  labs(
    x = "Station Name",
    y = "Usage",
  ) +
  scale_fill_manual(
    values = c("count" = "pink", "prediction" = "skyblue"),
    labels = c("count" = "Actual count in 2024", "prediction" = "Predicted count in 2025") 
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
